MIRSVM: Multi-Instance Support Vector Machine with Bag Representatives

### 2017-06-11 ArticleRevision/article/article.tex
Initial commit

\begin{abstract}
\textit{Multiple-instance learning} (MIL) is a generalization of \textit{supervised learning}, where samples are represented by labeled bags, each containing sets of instances. The individual labels of the samples within a bag are unknown, and labels are assigned based on a multi-instance assumption. One of the major complexities associated with this type of learning is the ambiguous relationship between a bag's label and the instances it contains. This paper proposes a novel support vector machine (SVM) multiple-instance formulation and presents an algorithm with a bag-representative selector that trains the SVM based on bag-level information, named MIRSVM. The contribution is able to identify instances that highly impact classification, i.e. bag-representatives, for both positive and negative bags, while finding the optimal class separation hyperplane. Unlike other multi-instance SVM methods, this approach eliminates possible class imbalance issues by allowing both positive and negative bags to have at most one representative, while highlighting the instances that contribute most to the classification model. The experimental study evaluates and compares the performance of this proposal against $10$ state-of-the-art multi-instance methods over $15$ datasets, and the results are validated through non-parametric statistical analysis. The results indicate that bag-based learners outperform the instance-based and wrapper methods, as well as this proposal's overall superior performance against other multi-instance SVM models. 
\keywords{Machine learning \and multiple-instance learning \and support vector machines \and bag-level multi-instance classification \and bag-representative selection}
\end{abstract}

\section{Introduction}\label{intro}
In traditional classification, learning algorithms attempt to correctly label unknown samples by finding patterns that exist between training samples and their class label. \textit{Multi-instance} (MI) learning (or \textit{multiple-instance} learning) is a generalization of \textit{supervised learning} that has been recently been gaining interest because of its applicability to many real-world problems such as text categorization \citep{Andrews2002}, image classification \citep{Herman2008,Wei2016155}, human action recognition \citep{Yi2016}, and drug activity prediction \citep{Dietterich1997}.

@@

\iitem Devising a unique bag-representative selector method that makes no presumptions about the underlying distributions of the instances within each bag, while maintaining the default MIL assumption. This approach eliminates the issue of class imbalance caused by techniques such as flattening or subsetting positive instances from each bag. The key feature of MIRSVM is its ability to identify instances (support vectors) within positive and negative bags that highly impact the model.
\end{itemize}

@@

\subsection{Notation}\label{subsec:notation}
Let $\mathcal{D}$ be a training dataset of $n$ bags. Let $\bm{Y} \in \mathcal{D}$ be a vector of $n$ labels corresponding to each bag, having a domain of $\bm{Y} \in \set{-1,\,+1}^n$. Let $\bm{X} \in \mathcal{D}$ be a matrix consisting of $d$ input variables and $m$ instances, having a domain of $\bm{X} \in \mathbb{R}^{m \times d}$. Let $\bm B$ be the set of bags which contain a number of instances, sometimes of different size, and usually non-overlapping, such as $\bm{B}_I = \{\bm x_{1}, \ldots, \bm x_{|\bm B_I|}\}$ for index set $I \subseteq \set{1,\ldots,n}$, where $|\bm B_I|$ is the number of instances in bag $I$. Using the training dataset $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$, the goal is to learn a model $f : \bm{X} \times Y$, that assigns a label, $Y_I \in \set{-1,\,+1}$, for each input bag $\bm{B}_I$. The model will then be used to predict the labels of new, unlabeled, and unseen input bags. Table~\ref{tab:Notation} provides a summary of the notation used in this paper.
\begin{table}
\caption{Summary of notation used throughout the paper}\label{tab:Notation}
\begin{tabularx}{\textwidth}{l@{\extracolsep{\fill}}l}
\hline\noalign{\smallskip}
Definition & Notation\\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
Number of Bags & $n$ \\
Number of Instances & $m$ \\
Number of Input Attributes & $d$ \\
Set of Bags & $\bm B = \set{\bm B_1,\, \ldots,\, \bm B_n}$ \\
Bag Index Set & $I \in \ints_+^n$ \\
Input Space & $\bm{X} \in \mathbb{R}^{m \times d}$ \\
Bag $I$ & $\bm B_I = \set{\bm x_i \st \forall i \in I}$ \\
Input Instance $i$ from Bag $I$ & $\bm{x}_i = (x_{1}, \ldots, x_{d}),\, i \in I$ \\
Bag Labels & $\bm{Y} \in \{-1,\,1\}^n$ \\
Individual Instance Label $i$ & $y_i \in \set{-1,+1}$ \\
\noalign{\smallskip}\hline\noalign{\smallskip}
Full Multi-Instance Training Dataset & $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$ \\
Full Single-Instance Training Dataset & $\mathcal{S} = \{(\bm x_1,y_1), \ldots, (\bm x_m,y_m)\} $\\
\noalign{\smallskip}\hline
\end{tabularx}
\end{table}

\subsection{Multiple-Instance Learning}\label{subsec:mil}
In traditional binary classification problems, the goal is to learn a model that maps input samples to labels, $f: \reals^{m \times d} \rightarrow Y^m \in \set{-1,\,+1}$. Multiple instance learning generalizes this framework and makes weaker assumptions about sample labeling. In the MIL case, samples are called \textit{bags} and each bag contains one or more input instances. Each bag is assigned a label, unlike traditional classification problems, where individual samples are assigned a label. Input instances, $\set{\bm x_1,\, \bm x_2,\, \ldots,\, \bm x_m}$, are grouped into bags with unique identifiers, $\bm B = \set{\bm B_1,\, \bm B_2,\, \ldots,\, \bm B_n \st \bm B_I = \set{\bm x_i \st \forall i \in I}}$, $\forall I \in \set{1, \ldots, n}$ and assigned a label, $Y_I$. An example representation of an MI dataset is shown in Figure~\ref{fig:mirep}.

@@

Instance-based methods that follow the SMI assumption attempt to identify desirable instance properties that make a bag positive. One traditional method in this category is the Axis-Parallel Rectangle (APR)~\citep{Dietterich1997}, which trains a model that assigns a positive label to an instance if it belongs to an axis-parallel rectangle in feature space, and assigns a negative label otherwise. The APR is optimized by maximizing the number of positive bags in the training set containing at least one instance in the APR, while concurrently maximizing the number of negative bags that do not contain any instance in the APR. Another similar method is the Diverse Density (DD)~\citep{Maron1998} framework which is maximized for instances in feature space that are near at least one instance in a positive bag and far from all instances in negative bags. In the Expectation-Maximization Diverse Density (EM-DD) algorithm,~\cite{Zhang2001} propose a similar framework that iteratively maximizes the DD measure.~\cite{Auer2004} present a boosting approach~\citep{Schapire19981651} that uses balls centered around positive bags to solve the MI problem called Multi-Instance Optimal Ball (MIOptimalBall). This approach is similar to that of APR and DD, except that~\cite{Auer2004} propose computing optimal balls per positive bags. A major challenge affecting these methods is that the distributions of the positive and negative bags affect their performance. Methods based on the DD metric~\citep{Chen2004,Chen2006,Carbonneau2016,Zhang2001} assume the positive instances form a cluster, which may not be the case. Alternatively~\cite{Fu2011} models the distribution of negative bags with Gaussian kernels, which can prove difficult when the quantity of data is limited. 
\begin{figure*}
\centering
\begin{tabular}{|c|ccccc|c|}
\hline
$\bm B$ & $F_1$ & $F_2$ & $F_3$ & \ldots & $F_d$ & $Y$ \\        
\hline
    	  & 0.1 & 0.8 & 2.5 & \ldots & 0.8 &    \\        
    $B_1$ 	  & 0.2 & 2.0 & 5.5 & \ldots & 3.0 & POS \\
    	  & 0.1 & ?   & 4.5 & \ldots & 0.1 &    \\
\hline
    	  & 1.5 & 4.0 & 0.8 & \ldots & 0.1 &    \\
    	  & 0.8 & 0.4 & 2.9 & \ldots & 1.1 &    \\
    $B_2$	  & 2.3 & 0.2 & 4.0 & \ldots & 5.5 & NEG \\
    	  & 6.7 & 5.0 & ?   & \ldots & 0.5 &    \\
    	  & 0.1 & 4.0 & 8.7 & \ldots & 3.3 &    \\
\hline
  \end{tabular}
  \caption{Example of multi-instance data representation. The figure shows a sample binary MI dataset with $2$ bags, identified as $B_1$ and $B_2$. The bags each contain differing number of instances, each with $d$ dimension, and are assigned positive and negative labels indicated by column $Y$. Missing instance feature values are denoted by a question mark.}%
  \label{fig:mirep}%
\end{figure*}

@@

\begin{figure}
    \centering
    \label{fig:diff}
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{mirsvm_figure.eps} % first figure itself
    \end{minipage}\hfill
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{misvm_figure.eps} % second figure itself
    \end{minipage}
    \caption{Difference between MIRSVM and MISVM on a random $2$-dimensional toy dataset. The instances of this dataset were drawn from the standard normal distribution, with $6$ bags, $4$ of which are negative (red) and $2$ of which are positive (blue), indicated by the red and blue dotted circles. There are a total of $125$ instances, $65$ belonging to the positive bags and $60$ belonging to the negative bags. The decision boundaries (black lines) are shown for each of the algorithms, along with the SVM margin (red and blue dashed lines) and support vectors (black crosses). In this example, both MIRSVM and MISVM were trained using $C = 1000$ and $\sigma = 1.5$. Note the differing number of support vectors produced by the two methods. MIRSVM has $6$, one for each bag, and MISVM has $29$. Also note the smoother representation of the data distribution given by MIRSVM's decision boundary, unlike MISVM whose decision boundary was greatly affected by the larger number of support vectors belonging to the negative class with respect to the only $2$ positive support vectors. }
\end{figure}
This formulation is designed to utilize and select representatives from positive and negative bags, unlike MISVM, which only optimizes over representatives from positive bags, while flattening the negative bag instances. MISVM allows multiple representatives to be chosen from negative bags and limits positive bag-representatives to be one, while MIRSVM allows for balanced bag-representative selection, where each bag is allowed one. MISVM also uses a wrapper technique to initialize the positive bag-representatives by taking the mean vector of the instances within each positive bag. This is an implicit assumption that the instances within the positive bags are all positive, whereas MIRSVM's initialization procedure selects an instance from all bags at random, ensuring no noise is added by any wrapper techniques during initialization and no assumptions are made about the instances. Due to the constraints on the representatives, MIRSVM produces sparser models while MISVM has the freedom to select as many negative support vectors as it needs and restricts the support vectors chosen from positive bags to be one. Figure~\ref{fig:diff} shows the decision boundaries produced by MIRSVM and MISVM to highlight the differences in their solutions. As Figure~\ref{fig:diff} shows, MISVM produces a higher number of support vectors from the negative bags, which greatly influences the final decision boundary in favor of the negative class.

>>>>

\begin{abstract}
\textit{Multiple-instance learning} (MIL) is a variation of \textit{supervised learning}, where samples are represented by labeled bags, each containing sets of instances. The individual labels of the samples within a bag are unknown, and labels are assigned based on a multi-instance assumption. One of the major complexities associated with this type of learning is the ambiguous relationship between a bag's label and the instances it contains. This paper proposes a novel support vector machine (SVM) multiple-instance formulation and presents an algorithm with a bag-representative selector that trains the SVM based on bag-level information, named MIRSVM. The contribution is able to identify instances that highly impact classification, i.e. bag-representatives, for both positive and negative bags, while finding the optimal class separation hyperplane. Unlike other multi-instance SVM methods, this approach eliminates possible class imbalance issues by allowing both positive and negative bags to have at most one representative, while highlighting the instances that contribute most to the classification model. The experimental study evaluates and compares the performance of this proposal against $10$ state-of-the-art multi-instance methods over $15$ datasets, and the results are validated through non-parametric statistical analysis. The results indicate that bag-based learners outperform the instance-based and wrapper methods, as well as this proposal's overall superior performance against other multi-instance SVM models. 
\keywords{Machine learning \and multiple-instance learning \and support vector machines \and bag-level multi-instance classification \and bag-representative selection}
\end{abstract}

\section{Introduction}\label{intro}
In traditional classification, learning algorithms attempt to correctly label unknown samples by finding patterns that exist between training samples and their class label. \textit{Multi-instance} (MI) learning (or \textit{multiple-instance} learning) is a variation of \textit{supervised learning} that has been recently been gaining interest because of its applicability to many real-world problems such as text categorization \citep{Andrews2002}, image classification \citep{Herman2008,Wei2016155}, human action recognition \citep{Yi2016}, and drug activity prediction \citep{Dietterich1997}.

@@

\iitem Devising a unique bag-representative selection method that makes no presumptions about the underlying distributions of the instances within each bag, while maintaining the default MIL assumption. This approach eliminates the issue of class imbalance caused by techniques such as flattening or subsetting positive instances from each bag. The key feature of MIRSVM is its ability to identify instances (support vectors) within positive and negative bags that highly impact the model.
\end{itemize}

@@

\subsection{Notation}\label{subsec:notation}
Let $\mathcal{D}$ be a training dataset of $n$ bags. Let $\bm{Y} \in \mathcal{D}$ be a vector of $n$ labels corresponding to each bag, having a domain of $\bm{Y} \in \set{-1,\,+1}^n$. Let $\bm{X} \in \mathcal{D}$ be a matrix consisting of $d$ input variables and $m$ instances, having a domain of $\bm{X} \in \mathbb{R}^{m \times d}$. Let $\mathcal{B}$ be the set of bags which contain a number of instances, sometimes of different size, and usually non-overlapping, such as $\bm{B}_I = \{\bm x_{1}, \ldots, \bm x_{|\bm B_I|}\}$ for index set $I \subseteq \set{1,\ldots,n}$, where $|\bm B_I|$ is the number of instances in bag $I$. Using the training dataset $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$, the goal is to learn a model $f : \bm{X} \times Y$, that assigns a label, $Y_I \in \set{-1,\,+1}$, for each input bag $\bm{B}_I$. The model will then be used to predict the labels of new, unlabeled, and unseen input bags. Table~\ref{tab:Notation} provides a summary of the notation used in this paper.
\begin{table}
\caption{Summary of notation used throughout the paper}\label{tab:Notation}
\begin{tabularx}{\textwidth}{l@{\extracolsep{\fill}}l}
\hline\noalign{\smallskip}
Definition & Notation\\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
Number of Bags & $n$ \\
Number of Instances & $m$ \\
Number of Input Attributes & $d$ \\
Set of Bags & $\mathcal{B} = \set{\bm B_1,\, \ldots,\, \bm B_n}$ \\
Bag Index Set & $I \in \ints_+^n$ \\
Input Space & $\bm{X} \in \mathbb{R}^{m \times d}$ \\
Bag $I$ & $\bm B_I = \set{\bm x_i \st \forall i \in I}$ \\
Input Instance $i$ from Bag $I$ & $\bm{x}_i = (x_{1}, \ldots, x_{d}),\, i \in I$ \\
Bag Labels & $\bm{Y} \in \{-1,\,1\}^n$ \\
Unknown Individual Instance Label $i$ & $y_i \in \set{-1,+1}$ \\
\noalign{\smallskip}\hline\noalign{\smallskip}
Full Multi-Instance Training Dataset & $\mathcal{D} = \{(\bm B_1,Y_1), \ldots, (\bm B_n,Y_n)\}$ \\
Full Single-Instance Training Dataset & $\mathcal{S} = \{(\bm x_1,y_1), \ldots, (\bm x_m,y_m)\} $\\
\noalign{\smallskip}\hline
\end{tabularx}
\end{table}

\subsection{Multiple-Instance Learning}\label{subsec:mil}
In traditional binary classification problems, the goal is to learn a model that maps input samples to labels, $f: \reals^{m \times d} \rightarrow Y^m \in \set{-1,\,+1}$. Multiple instance learning generalizes this framework and makes weaker assumptions about sample labeling. In the MIL case, samples are called \textit{bags} and each bag contains one or more input instances. Each bag is assigned a label, unlike traditional classification problems, where individual samples are assigned a label. Input instances, $\set{\bm x_1,\, \bm x_2,\, \ldots,\, \bm x_m}$, are grouped into bags with unique identifiers, $\mathcal{B} = \set{\bm B_1,\, \bm B_2,\, \ldots,\, \bm B_n \st \bm B_I = \set{\bm x_i \st \forall i \in I}}$, $\forall I \in \set{1, \ldots, n}$ and assigned a label, $Y_I$. An example representation of an MI dataset is shown in Figure~\ref{fig:mirep}.

@@

Instance-based methods that follow the SMI assumption attempt to identify desirable instance properties that make a bag positive. One traditional method in this category is the Axis-Parallel Rectangle (APR)~\citep{Dietterich1997}, which trains a model that assigns a positive label to an instance if it belongs to an axis-parallel rectangle in feature space, and assigns a negative label otherwise. The APR is optimized by maximizing the number of positive bags in the training set containing at least one instance in the APR, while concurrently maximizing the number of negative bags that do not contain any instance in the APR. Another similar method is the Diverse Density (DD)~\citep{Maron1998} framework which is maximized for instances in feature space that are near at least one instance in a positive bag and far from all instances in negative bags. In the Expectation-Maximization Diverse Density (EM-DD) algorithm,~\cite{Zhang2001} propose a similar framework that iteratively maximizes the DD measure.~\cite{Auer2004} present a boosting approach~\citep{Schapire19981651} that uses balls centered around positive bags to solve the MI problem called Multi-Instance Optimal Ball (MIOptimalBall). This approach is similar to that of APR and DD, except that~\cite{Auer2004} propose computing optimal balls per positive bags. A major challenge affecting these methods is that the distributions of the positive and negative bags affect their performance. Methods based on the DD metric~\citep{Chen2004,Chen2006,Carbonneau2016,Zhang2001} assume the positive instances form a cluster, which may not be the case. Alternatively~\cite{Fu2011} models the distribution of negative bags with Gaussian kernels, which can prove difficult when the quantity of data is limited. 
\begin{figure*}
\centering
\begin{tabular}{|c|ccccc|c|}
\hline
$\mathcal{B}$ & $F_1$ & $F_2$ & $F_3$ & \ldots & $F_d$ & $\bm Y$ \\        
\hline
    	  & 0.1 & 0.8 & 2.5 & \ldots & 0.8 &    \\        
    $B_1$ 	  & 0.2 & 2.0 & 5.5 & \ldots & 3.0 & POS \\
    	  & 0.1 & 0.1   & 4.5 & \ldots & 0.1 &    \\
\hline
    	  & 1.5 & 4.0 & 0.8 & \ldots & 0.1 &    \\
    	  & 0.8 & 0.4 & 2.9 & \ldots & 1.1 &    \\
    $B_2$	  & 2.3 & 0.2 & 4.0 & \ldots & 5.5 & NEG \\
    	  & 6.7 & 5.0 & 0.1  & \ldots & 0.5 &    \\
    	  & 0.1 & 4.0 & 8.7 & \ldots & 3.3 &    \\
\hline
  \end{tabular}
  \caption{Example of multi-instance data representation. The figure shows a sample binary MI dataset with $2$ bags, identified as $B_1$ and $B_2$. The bags each contain differing number of instances, each with $d$ dimension, and are assigned positive and negative labels indicated by column $\bm Y$.}%
  \label{fig:mirep}%
\end{figure*}

@@

\begin{figure}
    \centering
    \label{fig:diff}
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{mirsvm_figure.eps} % first figure itself
    \end{minipage}\hfill
    \begin{minipage}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{misvm_figure.eps} % second figure itself
    \end{minipage}
    \caption{Difference between MIRSVM and MISVM on a random $2$-dimensional toy dataset. The instances of this dataset were drawn from the standard normal distribution, with $6$ bags, $4$ of which are negative (red) and $2$ of which are positive (blue), indicated by the red and blue dotted circles. There are a total of $125$ instances, $65$ belonging to the positive bags and $60$ belonging to the negative bags. The decision boundaries (black lines) are shown for each of the algorithms, along with the SVM margin (red and blue dashed lines) and support vectors (black crosses). In this example, both MIRSVM and MISVM were trained using $C = 1000$ and $\sigma = 1.5$. Note the differing number of support vectors produced by the two methods. MIRSVM has $6$, one for each bag, and MISVM has $29$. Also note the smoother representation of the data distribution given by MIRSVM's decision boundary, unlike MISVM whose decision boundary was greatly affected by the larger number of support vectors belonging to the negative class with respect to the only $2$ positive support vectors. }
\end{figure}
This formulation is designed to utilize and select representatives from positive and negative bags, unlike MISVM, which only optimizes over representatives from positive bags, while flattening the negative bag instances. MISVM allows multiple representatives to be chosen from negative bags and limits positive bag-representatives to be one, while MIRSVM allows for balanced bag-representative selection, where each bag is allowed one. MISVM also uses a wrapper technique to initialize the positive bag-representatives by taking the mean vector of the instances within each positive bag. This is an implicit assumption that the instances within the positive bags are all positive, whereas MIRSVM's initialization procedure selects an instance from all bags at random, ensuring no noise is added by any wrapper techniques during initialization and no assumptions are made about the instances. Due to the constraints on the representatives, MIRSVM produces sparser models while MISVM has the freedom to select as many negative support vectors as it needs and restricts the support vectors chosen from positive bags to be one. Figure~\ref{fig:diff} shows the decision boundaries produced by MIRSVM and MISVM to highlight the differences in their solutions. As Figure~\ref{fig:diff} shows, MISVM produces a larger number of support vectors from the negative bags, which greatly influences the final decision boundary in favor of the negative class.
